If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-31 GroupPolarizationStatmod/paper/method.tex
updates to method

Our method combines several formal and computational tools to identify which
replications should be retracted and evaluate experimental designs across our
corpus. The first step in the analysis is to find a reasonable pair of latent
null-polarization distributions that appear polarized due to clipping.

\subsection{Identify replications for removal}

The first step is to find which replications of the 57 must be retracted 
due to non-identifiability. We establish non-identifiability by finding one
set of latent null-polarization opinion distributions where the mean doesn't change,
but appear polarized when measured on the ordinal scale used in the experiment.

To do
this we designed a web app that allows the user to try different latent
distribution parameters (Figure~\ref{fig:data-model-flow}) to seed a
hillclimbing algorithm that explores latent distribution parameter space
until it finds null parameters that look like group polarization under the
experiment's ordinal measurement procedure, or the algorithm stops after a
certain number of steps (Algorithm~\ref{alg:hillclimbing}). If the web app
finds

—latent mean and pre- and post-discussion 
variances that  loads experimental design and observation data
from a user-provided CSV file

ordinal measurements distort to look 
polarized.

To find examples of plausibly spurious group polarization we guess at what latent
mean, combined with a latent pre- and post-deliberation latent standard deviations,
could generate it. We calculate the observed group polarization for these latent
parameters by transforming the latent pre- and post-deliberation distributions from
continuous normal distributions to ordinal ``observations'' by integrating over each
ordinal bin, then calculating the observed mean for each experiment \emph{phase},
i.e., both the pre- and post-deliberation phases.  If the null model can generate
group polarization in this case, then the detection is labelled \emph{plausibly
false}, i.e., \emph{plausibly spurious}, and we call the generating latent
parameters the \emph{plausible null parameters}.

To evaluate whether the plausible null parameters we identify could realistically
fool a significance test, we implemented a generative simulation of each plausibly
false detection of group polarization and calculated the \emph{family-wise error rate
(FWER)} and the \emph{false discovery rate (FDR)}. We use the published number of
participants $N$ for each of the 54 conditions that yielded plausibly false
detections of group polarization. The FWER is also known as the Type I
error, or the false positive rate: it is the probability that a detection of an
effect occurs given that in reality there was a null effect.  The FDR is the
probability that a published positive detection of an effect is a false positive,
i.e., a detection of a null effect. High FWER and FDR for an experimental condition
indicates that its design is weak, since we identify only one of perhaps infinitely
many latent parameter combinations that can generate spurious group polarization.

@@

\subsection{Probability theory of spurious group polarization}

We formalize this experimental design in the $N\to\infty$ case where the
distribution of ordinal measurements is created by integrating continuous latent
opinion distributions.  Latent opinions in the model are normally distributed with
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  \omega_{i,t} \sim p(\omega; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent 
Latent opinions are assumed to be normally distributed since they are unbounded 
aggregations of neurobiological activity.

@@

\subsection{Interactive algorithm to find parameters inducing spurious group polarization}

@@

\subsection{Metascience of the false detection rate in group polarization}

Metascience provides a framework for
calculating the false discovery rate across experimental conditions,
aggregated to the level of journal articles and the ten-paper corpus. 
The false discovery rate is then the Type I error rate relative to the rate of
positive discoveries, true or false.
We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

@@

The group polarization studies are journal articles that present
the results of multiple experiments to support some hypothesis. To organize
the presentation of false detection rates across the ten studies in our
analysis, we first will organize false detection rates by experiment, where
we created `ExperimentID` tags for each of the sixty experiments across ten
studies. We also created `StudyID` tags for each of the ten journal articles
with the key `{Author}{Year}`, e.g., \texttt{Schkade2010} for Schkade, et al.,
(2010) \cite{Schkade2010}.

@@

\subsection{Estimating and limiting false detection rates}

>>>>

Our method combines several formal and computational tools to identify which
replications should be retracted and evaluate experimental designs across our
corpus. The first step in the analysis is to find a reasonable pair of latent
null-polarization distributions that appear polarized due to clipping
(Section~\ref{subsec:identify}). To do this, we develop a formal model and a numeric
rootfinding algorithm to power a Shiny web app we used to interactively identify and
record null-polarization parameters generating such measurement artifacts.

The second step in the analysis is to evaluate the quality of the experimental designs
when we properly account for the ordinal measurement procedure
(Section~\ref{subsec:evaluate}). We use the null-polarization parameters we found in
the first step to generate simulated experimental data, then estimate the difference
between the pre- and post-discussion means using ordered probit models to fit each
distribution—ordered probit models account for the ordinal measurement procedure.  We
estimate the false positive rate and false detection rate for low, medium, and high
significance thresholds in terms of Cohen's $d$. Finally, we use the effect sizes thus
calculated to determine the significance threshold one must use to achieve a 5\% false
positive rate.

\subsection{Identify replications for removal}
\label{subsec:identify}

The first step is to find which replications of the 57 must be retracted due to
non-identifiability. We establish non-identifiability by finding one set of latent
null-polarization opinion distributions where the mean doesn't change, but appear
polarized when measured on the ordinal scale used in the experiment.  In other words,
if we find a null-polarization model that explains the outcome as well as a group
polarization one, the group polarization hypothesis is \emph{non-identifiable} and the
replication must be rejected—the data cannot confirm or disconfirm the target
hypothesis.

To do this we designed a web app that allows the user to try different latent
distribution parameters (Figure~\ref{fig:data-model-flow}) to seed a
hillclimbing algorithm that explores latent distribution parameter space
until it finds null parameters that look like group polarization under the
experiment's ordinal measurement procedure, or the algorithm stops after a
certain number of steps (Algorithm~\ref{alg:hillclimbing}).

% To find examples of plausibly spurious group polarization we guess at what latent
% mean, combined with a latent pre- and post-deliberation latent standard deviations,
% could generate it. We calculate the observed group polarization for these latent
% parameters by transforming the latent pre- and post-deliberation distributions from
% continuous normal distributions to ordinal ``observations'' by integrating over each
% ordinal bin, then calculating the observed mean for each experiment \emph{phase},
% i.e., both the pre- and post-deliberation phases.  If the null model can generate
% group polarization in this case, then the detection is labelled \emph{plausibly
% false}, i.e., \emph{plausibly spurious}, and we call the generating latent
% parameters the \emph{plausible null parameters}.

% To evaluate whether the plausible null parameters we identify could realistically
% fool a significance test, we implemented a generative simulation of each plausibly
% false detection of group polarization and calculated the \emph{family-wise error rate
% (FWER)} and the \emph{false discovery rate (FDR)}. We use the published number of
% participants $N$ for each of the 54 conditions that yielded plausibly false
% detections of group polarization. The FWER is also known as the Type I
% error, or the false positive rate: it is the probability that a detection of an
% effect occurs given that in reality there was a null effect.  The FDR is the
% probability that a published positive detection of an effect is a false positive,
% i.e., a detection of a null effect. High FWER and FDR for an experimental condition
% indicates that its design is weak, since we identify only one of perhaps infinitely
% many latent parameter combinations that can generate spurious group polarization.

@@

\subsubsection{Probability theory of spurious group polarization}

% We formalize this experimental design in the $N\to\infty$ case where the
% distribution of ordinal measurements is created by integrating continuous latent
% opinion distributions.

Latent opinions are assumed to be normally distributed in general since they are
aggregations of many neurobiological potentials of
different polarities and magnitudes. We have two opinion distributions for 
two different times in the group polarization experiment process: pre-discussion,
$\tpre$, and post-discussion, $\tpost$.
The 
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  \omega_{i,t} \sim p(\omega; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent

@@

\subsubsection{Interactive algorithm to find parameters inducing spurious group polarization}

@@

\subsection{Evaluate experimental designs}

We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

@@

The false discovery rate is then the Type I error rate relative to the rate of
positive discoveries, true or false.
The group polarization studies are journal articles that present
the results of multiple experiments to support some hypothesis. To organize
the presentation of false detection rates across the ten studies in our
analysis, we first will organize false detection rates by experiment, where
we created `ExperimentID` tags for each of the sixty experiments across ten
studies. We also created `StudyID` tags for each of the ten journal articles
with the key `{Author}{Year}`, e.g., \texttt{Schkade2010} for Schkade, et al.,
(2010) \cite{Schkade2010}.

@@

\subsubsection{Estimating and limiting false detection rates}
